---
title: "GTEx comparison"
format: html
execute:
  message: false
  warning: false
---

```{r setup}
#| echo: false
#| message: false
library(readr)
library(dplyr)
library(duckdb)
library(googleCloudStorageR)
library(gt)
library(tidyr)
library(ggplot2)
```


This analysis compares the [eQTL Catalogue resources](https://github.com/eQTL-Catalogue/eQTL-Catalogue-resources)

The goal is to check what has changed between original release (used by the platform) and 
newly updated data from GTEx v10.

## Datasets

- data_tables
  - [original](https://github.com/eQTL-Catalogue/eQTL-Catalogue-resources/blob/master/data_tables/dataset_metadata.tsv)
  - [upcomming](https://github.com/eQTL-Catalogue/eQTL-Catalogue-resources/blob/master/data_tables/dataset_metadata_upcoming.tsv)
  
  
## Data loading

The analysis loads dataset tables from the `data_tables` mentioned above.

Note that *eQTL Catalogue* uses slightly different notion of a `study`, which
corresponds to a single `pubmed id`


```{r}
#| echo: false

file_paths <- list(
  original = "https://raw.githubusercontent.com/eQTL-Catalogue/eQTL-Catalogue-resources/refs/heads/master/data_tables/dataset_metadata.tsv", 
  upcoming = "https://raw.githubusercontent.com/eQTL-Catalogue/eQTL-Catalogue-resources/refs/heads/master/data_tables/dataset_metadata_upcoming.tsv")

```

```{r data-load}
#| echo: false
#| message: false
#| cache: true

data <- lapply(names(file_paths), function(name) {
  readr::read_tsv(file_paths[[name]]) |> dplyr::mutate(release = name)
}) |> 
  dplyr::bind_rows()|> 
  dplyr::filter(study_label == 'GTEx')
```


```{r}
#| echo: false
data |> 
  dplyr::distinct() |> 
  dplyr::group_by(study_id, pmid, release) |> 
  dplyr::summarise(n = n(), .groups = "drop") |>
  dplyr::arrange(study_id, pmid, release) |>
  gt::gt() |>
  gt::tab_header("Number of GTEx studies in releases")
```

The notion of a single Gentropy `study_id` is a concatenation of multiple values.
The `dataset_id` on the other hand represents a single row in the data_tables.
This is almost a proxy of a single `study_id`, but differs between the two 

Below you can see the number of `dataset_id(s)` that are repeated in both releases and addition of v10

```{r}
#| echo: false

dataset_id_counts <- data |>
  dplyr::group_by(dataset_id) |>
  dplyr::summarise(count = n())
  
single_release_ids <-  dataset_id_counts |> 
    dplyr::filter(count == 1) |> dplyr::pull(dataset_id)

single_release_datasets <- data |>
  dplyr::filter(dataset_id %in% single_release_ids) |>
  dplyr::select(dataset_id, release)

dataset_id_counts_with_releases <- dataset_id_counts |>
  left_join(single_release_datasets, by = "dataset_id") |>
  dplyr::mutate(
    count = factor(count),
    release = dplyr::case_when(
      count == 2 ~ "both",
      count == 1 ~ release,
      .default = as.character(count)
    )
  )
  
  
ggplot2::ggplot(
  data = dataset_id_counts_with_releases, 
  mapping = ggplot2::aes(count, fill = release)
) + 
  ggplot2::geom_bar(color = "black", linewidth = 0.5) + 
  ggplot2::theme_minimal()

  
```


### Comparison of the datasets

```{r}
#| echo: false

sample_sizes_per_release_common_ds <- data |> 
  dplyr::filter(!dataset_id %in% single_release_ids) |>
  dplyr::mutate(
    release = dplyr::case_when(
      release == "upcoming" ~ "upcoming(both)",
      release == "original" ~ "original(both)"
    )
  ) |>
  dplyr::union_all(data |> 
    dplyr::filter(dataset_id %in% single_release_ids))

ggplot2::ggplot(
  data = sample_sizes_per_release_common_ds,
  mapping = ggplot2::aes(x = release, y = sample_size)) +
  ggplot2::geom_boxplot(alpha = 0.7) +
  ggplot2::stat_summary(
    fun = max, 
    geom = "text", 
    mapping = ggplot2::aes(label = ggplot2::after_stat(y)),
    vjust = -0.5, color = "red", size = 3, fontface = "bold") +
  ggplot2::stat_summary(
    fun = min, 
    geom = "text", 
    mapping = ggplot2::aes(label = ggplot2::after_stat(y)),
    vjust = 1.5, color = "blue", size = 3, fontface = "bold") +
  ggplot2::stat_summary(
    fun = median, 
    geom = "text", 
    mapping = ggplot2::aes(label = ggplot2::after_stat(y)),
     vjust = -0.5, color = "darkgreen", size = 3, fontface = "bold") +
  ggplot2::labs(
    title = "Sample Sizes by Release Type",
    x = "Release",
    y = "Sample Size"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)
  )

```

Based on the data from both releases it looks like, the `upcoming` release has relatively larger `sample_size` in all datasets.

The `both` - common datasets in both releases are different only in the `sample_size`.

::: {.callout-note collapse="true"}
### Check exact differences
```{r}
#| echo: false
# show example data rows that have the same study_id and dataset_id
data |> 
  dplyr::filter(!dataset_id %in% single_release_ids) |>
  dplyr::arrange(dataset_id) |>
  gt::gt() |> 
  gt::tab_style(
    style = cell_fill(color = "lightyellow"),
    locations = cells_body(columns = sample_size)
  )
```
:::

### quantification method

The new `upcoming` datasets are introducing the new `quant_method` - **majiq**

```{r}
#| echo: false


data |> 
  dplyr::filter(!dataset_id %in% single_release_ids) |>
  dplyr::mutate(
    release = dplyr::case_when(
      release == "upcoming" ~ "upcoming(both)",
      release == "original" ~ "original(both)"
    )
  ) |>
  dplyr::union_all(data |> 
    dplyr::filter(dataset_id %in% single_release_ids)) |>
  dplyr::group_by(quant_method, release) |>
  dplyr::summarise(cnt = n(), .groups = "drop") |> 
  dplyr::arrange(release, quant_method) |>
  gt::gt() |>  
  gt::tab_style(
    style = cell_fill(color = "lightyellow"),
    locations = cells_body(rows = quant_method == "majiq")
  )

```
  
### Single-cell vs bulk

All new datasets are from bulk sequencing.

```{r}
#| echo: false


data |> 
  dplyr::filter(!dataset_id %in% single_release_ids) |>
  dplyr::mutate(
    release = dplyr::case_when(
      release == "upcoming" ~ "upcoming(both)",
      release == "original" ~ "original(both)"
    )
  ) |>
  dplyr::union_all(data |> 
    dplyr::filter(dataset_id %in% single_release_ids)) |>
  dplyr::group_by(study_type, release) |>
  dplyr::summarise(cnt = n(), .groups = "drop") |> 
  dplyr::arrange(release, study_type) |>
  gt::gt() |>  
  gt::tab_style(
    style = cell_fill(color = "lightyellow"),
    locations = cells_body(rows = release == "upcoming")
  )

```
  

## Population assignments

Below is the population assignment from [population_assignment file](https://github.com/eQTL-Catalogue/eQTL-Catalogue-resources/blob/master/data_tables/population_assignments.tsv)

```{r}
#| echo: false
#| message: false

pops = readr::read_tsv("https://raw.githubusercontent.com/eQTL-Catalogue/eQTL-Catalogue-resources/refs/heads/master/data_tables/population_assignments.tsv") |>
  dplyr::filter(Study == "GTEx")

pops |>
  gt::gt() |> 
  gt::tab_style(
    style = cell_fill(color = "lightyellow"),
    locations = cells_body(rows = Study == "GTEx")
  )

```

Note: the file is not updated and seem to represent **total number of participants**, which
does not match the `sample_size` numbers in the individual datasets, **thus it can not be used
to infer the LD population structure.**


## Summary

With update we should expect to:
- Deprecate 245 studies
- Introduce 294 (245 old + 49 new) studies with increased sample_size
- New 49 studies are introducing new quantification method - **majiq**




